Foundational claims of group polarization are plausibly false, with lessons for the empirical study of psychological dynamics

### 2024-11-01 GroupPolarizationStatmod/paper/ms.tex
cleaning up for submission-ready and starting FDR calcs

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=1.1in,letterpaper]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

@@

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\begin{quote}
In our introductory social psychology course, 
we have for many years used the [group polarization experimental paradigm] as
a laboratory exercise. The exercise works beautifully, but one must be
careful to forewarn a class that [group polarization] does not occur with every 
group\ldots and that the effect is not large. 
\par\raggedleft(Brown, 1986, p.\cite[p. 205]{Brown1986})
\end{quote}

@@

@@

Unchecked social and political polarization undermines cooperation and
political efficacy \autocite{Mason2018UncivilAgreementBook,Klein2020}.
\emph{Group polarization} is the name for a group-level process that
leads to greater opinion extremity in a group as the group deliberates
some topic (e.g., ``Do you approve, disapprove, or neither approve nor
disapprove of the current administration's policies?''). Extremity only
increases, though, assuming that the mean opinion of the group is
initially biased towards one extreme or the other
\autocite{Brown1986,Brown2020}. Group polarization is important to
understand rigorously, therefore, since if groups become increasingly
extreme over time, while group opinions are consolidated, this would
exacerbate political polarization between political opposition groups,
which is already undermines the ability of partisans in democratic
governments to work together. Dozens of studies and meta-analyses over
decades have apparently established the empirical reality of group
polarization
\autocite{Moscovici1969,Myers1970,Isenberg1986,Sunstein2009,Sunstein2019}
\autocite[Ch. 5]{Brown2000}. We believe group polarization should occur
in certain contexts since more extreme group members are likely also to
be more stubborn \autocite{Guazzini2015,Lewandowsky2019}, which should
pull consensus to extremes \autocite{Turner2020}. However, many or most
of the evidence for group polarization relies on experimental designs
that use ordinal measurements that are then analyzed as if they were
continuous. Unfortunately, this procedure is unfortunately known to
induce false positives if it is not properly accounted for
\autocite{Liddell2018}.

@@

We begin by developing a formal model of ordinal measurement of latent
continuous opinions (model variables and parameters are given in
Table~\ref{tab:modelVariables}). First, we assume that participant $i$ has
latent opinion at time $t$ that is drawn from a group-level normal distribution,
\begin{equation}
  o_{i,t} \sim p(o; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution}
\end{equation} 
\noindent 
The initial opinion of agent $i$ is are denoted by $o_{i,t=0}$, and thus
in the model are drawn from a normal distribution with mean $\mu_0$ and standard
deviation $\sigma_0$. The final opinion of agent $i$ is written $o_{i,t=T}$,
drawn from a normal distribution with mean $\mu_T$ and standard deviation
$\sigma_T$.

@@

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:MetricBoxplot} ABOUT HERE ]} \\
\end{center}

\vspace{0.5em}
\begin{center}
{[ FIGURE \ref{fig:OrdinalBoxplot} ABOUT HERE ]} \\
\end{center}

@@

\section{Figures}

\begin{figure}
  \caption{
    Boxplots visualizing the distribution of Cohen's $d$ values (x-axis in
    A and B) inferred through a metric linear model over 1000 simulation 
    trials for each experimental condition (y-axis in A and B). All conditions
    have many trials that result in non-zero Cohen's $d$ values with number of
    simulated participants, $N$, set to their empirical values reported in the
    associated journal article (A). However, when the number of simulated
    participants is increased to $10N$, there are several conditions that 
    indeed yield distributions closer to $d = 0$ as expected, however some
    conditions have $d$ distributions centered on a nonzero mean with little to
    no overlap with $d=0$; variance in $d$ decreases as expected
    across experimental conditions (B).
  }  
  \label{fig:MetricBoxplots}
  \centering
  \begin{subfigure}[]{0.9\textwidth}
    % \hspace{-7.5em} 
    \centering
    \textsf{\large \textbf{A}, empirical $N$} \\
    \includegraphics[width=\textwidth]{Figures/boxplots/metric_cohens.pdf}
    \label{fig:MetricBoxplot_N}
  \end{subfigure} \\
\end{figure}
\begin{figure}\ContinuedFloat
  \begin{subfigure}[]{0.9\textwidth}
    % \hspace{-5.5em} 
    \centering
    \textsf{\large \textbf{B}, 10 $\times$ empirical $N$} \\
    \includegraphics[width=\textwidth]{Figures/boxplots/10N_metric_cohens.pdf}
    \label{fig:MetricBoxplot_10N}
  \end{subfigure}
\end{figure}

\begin{figure}
  \caption{Boxplot of Cohen's $d$ for the ordered probit model over 100
    simulation trias for each experimental condition (y-axis in A and B). Each
    trial represents a possible outcome of a group polarization experiment where
    the true opinion shift is zero. The mean across trials is closer to 0 than
    in the metric case, but many simulated zero-shift experiments still result in
    an observed shift. This illustrates the weak statistical power of these group
    polarization experiments.}
  \label{fig:OrdinalBoxplot}
  \centering
    \includegraphics[width=0.9\textwidth]{Figures/boxplots/ordinal_cohens.pdf}
\end{figure}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=1.1in,letterpaper]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\newcommand{\mupre}{\mu_\mathrm{pre}}
\newcommand{\mupost}{\mu_\mathrm{post}}

\newcommand{\sigmapre}{\sigma_\mathrm{pre}}
\newcommand{\sigmapost}{\sigma_\mathrm{post}}

\newcommand{\thetafit}{\theta_\mathrm{fit}}

\newcommand{\fdr}{\mathrm{FDR}}

\definecolor{myorange}{RGB}{240, 96, 0}
\newcommand{\mt}[1]{{\textcolor{myorange} {({\tiny MT:} #1)}}}

@@

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=gray}
\setcounter{tocdepth}{3}
\tableofcontents
}

\begin{center} \noindent\rule{4cm}{0.4pt} \end{center}
\clearpage

\begin{quote}
In our introductory social psychology course, 
we have for many years used the [group polarization experimental paradigm] as
a laboratory exercise. The exercise works beautifully, but one must be
careful to forewarn a class that [group polarization] does not occur with every 
group\ldots and that the effect is not large. 
\par\raggedleft(Brown, 1986, p.\cite[p. 205]{Brown1986})
\end{quote}

@@

\begin{quote}
A medium effect size is conceived as one large enough to be visible to the naked eye.
\par\raggedleft(Cohen, 1988 \cite[p. 26]{CohenBook1988})
\end{quote}

@@

the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.

Unchecked social and political polarization undermines cooperation and political
efficacy \autocite{Mason2018UncivilAgreementBook,Klein2020}.  \emph{Group
polarization} is the name that has been given to a group-level process that leads
to greater opinion extremity in a group as the group deliberates some topic (e.g.,
``Do you approve, disapprove, or neither approve nor disapprove of the current
administration's policies?''). Extremity only increases, though, assuming that the
mean opinion of the group is initially biased towards one extreme or the other
\autocite{Brown1986,Brown2020}. Group polarization is important to understand
rigorously, therefore, since if groups become increasingly extreme over time,
while group opinions are consolidated, this would exacerbate political
polarization between political opposition groups, which is already undermines the
ability of partisans in democratic governments to work together. Dozens of studies
and meta-analyses over decades have apparently established the empirical reality
of group polarization
\autocite{Moscovici1969,Myers1970,Isenberg1986,Sunstein2009,Sunstein2019}
\autocite[Ch. 5]{Brown2000}. We believe group polarization should occur in certain
contexts since more extreme group members are likely also to be more stubborn
\autocite{Guazzini2015,Lewandowsky2019}, which should pull consensus to extremes
\autocite{Turner2020}. However, many or most of the evidence for group
polarization relies on experimental designs that use ordinal measurements that are
then analyzed as if they were continuous. Unfortunately, this procedure is
unfortunately known to induce false positives if it is not properly accounted for
\autocite{Liddell2018}.

@@

% We begin by developing a formal model of ordinal measurement of latent
% continuous opinions (model variables and parameters are given in
% Table~\ref{tab:modelVariables}).

Metascience frames our analysis since metascientific theory provides a framework for
our goal of calculating the false discovery rate across experimental conditions,
journal articles, and the ten-paper corpus. The false discovery rate is then
the Type I error rate relative to the rate of positive discoveries, true or false.
We calculate via simulation and data fitting
the Type I error rate for the experimental designs used in experimental conditions
that, under a preliminary large-$N$ model analysis, generated group
polarization upon ordinal measurement of simple consensus latent opinion 
distributions. These can then be aggregated to estimate the false discovery rate
for each journal article and for the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.

\subsection{Metascience of the false detection rate}

The false detection rate is the scaled probability that a detection, $D$, of
an effect, $E$, is in fact a false detection and there is \emph{not} an effect,
denoted $\neg E$. The probability that a detection $D$ has occurred when there
was no effect, i.e. when $\neg E$, is denoted $\alpha = \Pr(D | \neg E)$.
This is known as the Type I or family-wise error rate. To calculate the 
false detection rate ($\fdr$), one must compare $\alpha$ to the statistical power,
the probability of a detection when there is an effect, denoted $W = \Pr(D | E)$.
Both these values must be scaled by the rate at which effects occur, denoted $b$.
Using the notation outlined above, the false detection rate is
\begin{equation}
  \fdr = \frac{(1 - b)\alpha}{(1 - b)\alpha + bW}.
\end{equation}

We calculate $\alpha = \Pr(D | \neg E)$ through simulation for experimental
conditions identified as plausibly generating false detections, 
and use generously optimistic values of
$b=0.1$~\cite{Ioannidis2005,McElreath2015} and $W=0.8$~\cite{Smaldino2016NatSel}
in the main text and in the compare with more pessimistic, yet realistic, estimates
of these values, as well as even more optimistic estimates.

In group polarization, $E$ represents the group polarization \emph{effect}, and
$\neg E$ represents its absence. Simple consensus is one way for group polarization
to be absent, i.e., simple consensus is one form of $\neg E$. The presence of
an effect $E$ is independent of whether an effect was \emph{detected}, 
represented by $D$ if it was detected and $\neg D$ if not.

\subsection{How study design can induce false detections of group polarization}

First, we assume that participant $i$ has
latent opinion at time $t$ that is drawn from a group-level normal distribution,
\begin{equation}
  o_{i,t} \sim p(o; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution}
\end{equation} 
\noindent 
The initial opinion of agent $i$ is are denoted by $o_{i,t=0}$, and thus
in the model are drawn from a normal distribution with mean $\mu_0$ and standard
deviation $\sigma_0$. The final opinion of agent $i$ is written $o_{i,t=T}$,
drawn from a normal distribution with mean $\mu_T$ and standard deviation
$\sigma_T$.

\subsection{Root-finding algorithm to identify plausible false detections}

Simple consensus can appear as group polarization since a change in

\begin{algorithm}
  \caption{User-guided routine to find pre- and post-deliberation variances that
  yield group polarization in the absence of mean opinion change.
\mt{NOT YET DONE...}}\label{alg:routine}
  \begin{algorithmic}
    \Require $\mu$, $\sigmapre$, $\sigmapost$ \Comment{Static guess for common mean,
    initial guesses of variances.}
    \Require $\epsilon$, $\delta$, $i_\mathrm{max}$ \Comment{Root-finding tolerance, step size, and maximum
    iterations.}
    % \Require $\delta$ \Comment{Root-finding step size.}
    % \Require $i_\mathrm{max}$ \Comment{Maximum number of iterations.}
    \State $\Delta \gets \mupre' - \mupost'$ 
    \While {$\epsilon < |\Delta| $ \and $i < i_max$}
      \State $\Delta \gets (\mupre' - \mupost')^2$
    \EndWhile
  \end{algorithmic}
\end{algorithm}

\subsection{Calculate false detection rate from simulations}

False detection rates are calculated using Cohen's $d$ to measure effect
size, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
\end{equation}. 
\noindent
Denote the collection of fitted parameters from the $i^\mathrm{th}$ simulation trial
for experimental condition $c$ and journal article $a$ as
% $\thetafit = \left{\mu_t = \left{\mupre,~\mupost\right},~\sigma_t = \left{\mupre,~\mupost\right}\right}$.
$\thetafit_{a,c,i} = \{\mupre,~\mupost,~\mupre,~\mupost\}_{a,c,i}$.
These parameters are the modal estimates for each parameter obtained from
an ordered probit statistical model fit with
Bayesian Markov-Chain Monte Carlo (MCMC)~\cite[p. 338]{Liddell2018}.

For each $a,c$, we calculate $\alpha_{a,c}(d^*) =
\Pr(D |\neg E;~a,~c,~d^*)$ for a given significance value, $d^*$, directly as
the fraction of $d$ values that are greater than $d^*$.

To calculate article-level false detection curves we calculate the average
across , we set $\alpha_{a,c} = 0.0$ for those
conditions for which we failed to generate spurious group polarization in the
large-$N$ model.

@@

\subsection{Ninety percent of detections are plausibly false}

We identified 54 of 60 experimental 
conditions where a $N \to \infty$ model falsely indicated group polarization 
instead of simple consensus.

\subsection{Eighty percent false detection rate across corpus for $d^*=0.8$?}

\subsection{What $d^*$ is necessary to achieve magical $\alpha = 0.05$?}

@@

% \section{Figures}

% \begin{figure}
%   \caption{
%     Boxplots visualizing the distribution of Cohen's $d$ values (x-axis in
%     A and B) inferred through a metric linear model over 1000 simulation 
%     trials for each experimental condition (y-axis in A and B). All conditions
%     have many trials that result in non-zero Cohen's $d$ values with number of
%     simulated participants, $N$, set to their empirical values reported in the
%     associated journal article (A). However, when the number of simulated
%     participants is increased to $10N$, there are several conditions that 
%     indeed yield distributions closer to $d = 0$ as expected, however some
%     conditions have $d$ distributions centered on a nonzero mean with little to
%     no overlap with $d=0$; variance in $d$ decreases as expected
%     across experimental conditions (B).
%   }  
%   \label{fig:MetricBoxplots}
%   \centering
%   \begin{subfigure}[]{0.9\textwidth}
%     % \hspace{-7.5em} 
%     \centering
%     \textsf{\large \textbf{A}, empirical $N$} \\
%     \includegraphics[width=\textwidth]{Figures/boxplots/metric_cohens.pdf}
%     \label{fig:MetricBoxplot_N}
%   \end{subfigure} \\
% \end{figure}
% \begin{figure}\ContinuedFloat
%   \begin{subfigure}[]{0.9\textwidth}
%     % \hspace{-5.5em} 
%     \centering
%     \textsf{\large \textbf{B}, 10 $\times$ empirical $N$} \\
%     \includegraphics[width=\textwidth]{Figures/boxplots/10N_metric_cohens.pdf}
%     \label{fig:MetricBoxplot_10N}
%   \end{subfigure}
% \end{figure}

% \begin{figure}
%   \caption{Boxplot of Cohen's $d$ for the ordered probit model over 100
%     simulation trias for each experimental condition (y-axis in A and B). Each
%     trial represents a possible outcome of a group polarization experiment where
%     the true opinion shift is zero. The mean across trials is closer to 0 than
%     in the metric case, but many simulated zero-shift experiments still result in
%     an observed shift. This illustrates the weak statistical power of these group
%     polarization experiments.}
%   \label{fig:OrdinalBoxplot}
%   \centering
%     \includegraphics[width=0.9\textwidth]{Figures/boxplots/ordinal_cohens.pdf}
% \end{figure}
